---
title: "Zebrafish biogeography by sex: Diversity analysis"
output: html_notebook
---

### Set Environment

```{r}

library(dada2)
library(ggplot2)
library(vegan)
library(reshape2)
library(glmmTMB)
library(MASS)
options(stringsAsFactors = F)

```

### Import data

```{r}

seqtab.nochim <- read.table("../../data/dada2_out/seqtab_nochim.txt",
                            header = T,
                            row.names = 1,
                            sep = "\t"
                          )

taxa <- read.table("../../data/dada2_out/taxa.txt",
                            header = T,
                            row.names = 1,
                            sep = "\t"
                          )

#make metadata
females <- c("F1", "F4", "F5", "F8", "F12") 

metadata <- data.frame(row.names = rownames(seqtab.nochim),
                       full_name = rownames(seqtab.nochim), 
                       id = sapply(strsplit(rownames(seqtab.nochim), "_"), 
                                   FUN = function(x){x[1]}), 
                       tissue = sapply(strsplit(rownames(seqtab.nochim), "_"), 
                                       FUN = function(x){x[2]})
                       )

#add sex : males = 0 Female = 1 
metadata$sex <-  as.numeric(metadata$id %in% females)

#factor tissue with the right levels 
metadata$tissue <- factor(metadata$tissue,
       levels = c("PG", "MG", "HG", "F")
       )

metadata$sex <- factor(metadata$sex)

write.table(metadata,
            file = "../../data/metadata_biogeo.txt",
            sep = "\t",
            col.names = T,
            row.names = T,
            quote = F
            )

```



### Normalize

```{r}

raresamp <- min(rowSums(seqtab.nochim))

#now we can plot the curves
rarecurve(seqtab.nochim, step = 1000,
          sample = 0,
          label = F,
          col = "black")

abline(v = 1000, col = "red")
abline(v = 2500, col = "yellow")
abline(v = 5000, col = "black")
abline(v = 10000, col = "black")


set.seed(731)
seqtab.nochim.rare <- rrarefy(seqtab.nochim, 10000) #rarefy
seqtab.nochim.rare <- seqtab.nochim.rare[,which(colSums(seqtab.nochim.rare) > 0)]

seqtab.nochim.rclr <- decostand(seqtab.nochim, method = "rclr") #center log blah blah blah
seqtab.nochim.rel <- decostand(seqtab.nochim, method = "total") #relative abundance

write.table(t(seqtab.nochim.rare),
            file = "../../data/rarefied_fish_biogeo.txt",
            sep = "\t",
            col.names = T,
            row.names = T,
            quote = F
            )

```


### Phylotyping

First we need a function that can aggregate asv counts data


```{r tax_functions}

###
#        Function phylotype_analysis             #
###

phylotype_analysis <- function(obj, tax) {
  #obj: microbiome object with at least 1 slot (data)
  #tax: a tax object (named list taxa as names values in the list are seq ids)
  obj.out <- NULL
  for (h in 1:length(tax)) {
    df <- NULL
    #print(h)#debugging
    for (i in 1:length(tax[[h]])) {
      #print(i)#debugging
      v1       <- obj$data[, unlist(tax[[h]][[i]])]
      v2       <- names(tax[[h]])[i]
      if (is.null(dim(v1))) {
        df[[v2]] <- v1
      } else{
        df[[v2]] <- rowSums(v1)
      }
    }
    obj.out[[names(tax)[h]]] <- as.data.frame(df)
  }
  return(obj.out)
}

make_taxa_df <- function(tax){

  kingdom.df <- replicate(length(unique(tax[, 2])), c())
  names(kingdom.df) <- unique(tax[, 2])
  phylum.df  <- replicate(length(unique(tax[, 3])), c())
  names(phylum.df) <- unique(tax[, 3])
  class.df   <- replicate(length(unique(tax[, 4])), c())
  names(class.df) <- unique(tax[, 4])
  order.df   <- replicate(length(unique(tax[, 5])), c())
  names(order.df) <- unique(tax[, 5])
  family.df  <- replicate(length(unique(tax[, 6])), c())
  names(family.df) <- unique(tax[, 6])
  genus.df   <- replicate(length(unique(tax[, 7])), c())
  names(genus.df) <- unique(tax[, 7])

  for (i in 1:nrow(tax)) {
    kingdom.df[[tax[i, 2]]] <-
      c(kingdom.df[[tax[i, 2]]], tax[i, 1])
    phylum.df[[tax[i, 3]]]  <-
      c(phylum.df[[tax[i, 3]]], tax[i, 1])
    class.df[[tax[i, 4]]]   <-
      c(class.df[[tax[i, 4]]], tax[i, 1])
    order.df[[tax[i, 5]]]   <-
      c(order.df[[tax[i, 5]]], tax[i, 1])
    family.df[[tax[i, 6]]]  <-
      c(family.df[[tax[i, 6]]], tax[i, 1])
    genus.df[[tax[i, 7]]]   <-
      c(genus.df[[tax[i, 7]]], tax[i, 1])
  }

  tax.obj <- NULL
  tax.obj$kingdom <- kingdom.df
  tax.obj$phylum  <- phylum.df
  tax.obj$class   <- class.df
  tax.obj$order   <- order.df
  tax.obj$family  <- family.df
  tax.obj$genus   <- genus.df

  return(tax.obj)
}

```

Now to make taxon level tables we simply apply the functions. Note, if we are using a filtered table (i.e., rarefied table) some ASVs may be inconsistent between the original (nochim) and the filtered table which will cause problems for the above functions. We can still use these tables, but we must filter as below first

```{r taxon_table_generation}

#we need to add an ASV identifier for the tax function to work properly
#we also need to restrict this analysis to only ASV IDs present in the data
#we will generate a obj in the process (below)

tax.bio <- taxa[which(rownames(taxa) %in% colnames(seqtab.nochim.rare)),]
tax.bio <- data.frame(asv = rownames(tax.bio), tax.bio)

#make a data frame of taxonomy
biogeo.tax <- make_taxa_df(tax = tax.bio)

# aggregate phylotype counts for this we will need to provide an object with a
# slot called data
biogeo.obj <- NULL
biogeo.obj$data <- seqtab.nochim.rare

#classifying taxonomy
biogeo_taxonomy <- phylotype_analysis(biogeo.obj, tax = biogeo.tax)

```


### Alpha diversity 

#### Calculate Richness and Entropy

```{r}

#richness 
richness.rare <- data.frame(specnumber(seqtab.nochim.rare))
colnames(richness.rare) <- "richness"
head(richness.rare)

#shannon entropy
shannon.rare <- data.frame(diversity(seqtab.nochim.rare, index = "shannon"))
colnames(shannon.rare) <- "shannon"
head(shannon.rare)

richness.fit <- glmmTMB(richness.rare$richness ~ tissue + sex + (1|id), data = metadata)

shannon.fit <- glmmTMB(shannon.rare$shannon ~ tissue + sex + (1|id), data = metadata)

summary(richness.fit)
summary(shannon.fit)

Anova(richness.fit)
Anova(shannon.fit)

```


#### Plot Richness and Entropy

```{r}

richness.rare$tissue <- metadata$tissue
richness.rare$sex <- metadata$sex

shannon.rare$tissue <- metadata$tissue
shannon.rare$sex <- factor(metadata$sex)



rich.plot <- ggplot(data = richness.rare, aes(x = tissue,
                                              y = richness,
                                              fill = factor(sex)))  

png(filename = "../../analysis/figs/sex_biogeo_richness.png", res = 300, width = 7,height = 7, units = "in")

rich.plot + 
  geom_boxplot()+
  scale_fill_manual("Sex", values = c("#ffe8d6", "#495877"), labels = c("Male", "Female"))+
  scale_x_discrete("", labels =c ("PG"="Proximal", "MG"="Middle", "HG"="Distal","F"="Feces"))+
 
  theme(
    text = element_text(size = 16, colour = "black"),
    panel.grid.major = element_line(colour = "grey99"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    aspect.ratio = 1,
  ) + 
  ylab("Richness")

dev.off()


shannon.rare$tissue <- metadata$tissue
shannon.rare$sex <- metadata$sex

shannon.plot <- ggplot(data = shannon.rare, aes(x = tissue,
                                                y = shannon,
                                                fill = factor(sex))) 

png(filename = "../../analysis/figs/sex_biogeo_shannon.png", res = 300, width = 7,height = 7, units = "in")

shannon.plot + 
  geom_boxplot()+
  scale_fill_manual("Sex", values = c("#ffe8d6", "#495877"), labels = c("Male", "Female"))+
  scale_x_discrete("", labels =c ("PG"="Proximal", "MG"="Middle", "HG"="Distal","F"="Feces"))+
 
  theme(
    text = element_text(size = 16, colour = "black"),
    panel.grid.major = element_line(colour = "grey99"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    aspect.ratio = 1,
  ) + 
  ylab("Shannon")

dev.off()


```


### Beta Diversity


```{r rel_abund}

seq.rare.prcomp <- prcomp(seqtab.nochim.rel, scale =F, center = F) # defaults
summary(seq.rare.prcomp)
plot(seq.rare.prcomp)

seqtab_rare_pca.df <- as.data.frame(seq.rare.prcomp$x[,1:5])

#add metadata for plotting
seqtab_rare_pca.df$tissue <- metadata$tissue
seqtab_rare_pca.df$sex <- metadata$sex
seqtab_rare_pca.df$id <- metadata$id


seqtab_rare_pca.plot <- ggplot(data = seqtab_rare_pca.df,
                                 aes(x = PC1,
                                     y = PC2,
                                     color = tissue, 
                                     shape = factor(sex)
                                     ) 
                               ) 
seqtab_rare_pca.plot +
  geom_point(size = 3, alpha = .7)+
  scale_color_brewer(palette = "Dark2")+
  theme(
    text = element_text(size = 16, colour = "black"),
    panel.grid.major = element_line(colour = "grey99"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    aspect.ratio = 1,
  )+
  xlab("PC1 (63%)")+
  ylab("PC2 (34%)")

```

We can try something similar with the log ratio transformed too, but it wont work with PCA like I originally thought.

```{r robust_aitchison}
#PCoA 
seq.rclr.cmdscale <- cmdscale(vegdist(x = seqtab.nochim, method = "robust.aitchison"), k =5, eig = T)

#percent variation explained 
round(seq.rclr.cmdscale$eig*100/sum(seq.rclr.cmdscale$eig),1)

#grab the dims
seq.rclr.cmdscale.df <- as.data.frame(seq.rclr.cmdscale$points)

#add metadata for plotting
seq.rclr.cmdscale.df$tissue <- metadata$tissue
seq.rclr.cmdscale.df$sex <- metadata$sex
seq.rclr.cmdscale.df$id <- metadata$id

colnames(seq.rclr.cmdscale.df) <- c("Dim1","Dim2", "Dim3", "Dim4", "Dim5", "tissue", "sex", "id")


seq.rclr.cmdscale.plot <- ggplot(data = seq.rclr.cmdscale.df,
                                 aes(x = Dim1,
                                     y = Dim2,
                                     color = tissue, 
                                     shape = factor(sex)
                                     ) 
                               ) 
seq.rclr.cmdscale.plot +
  geom_point(size = 3, alpha = .7)+
  scale_color_brewer(palette = "Dark2")+
  theme(
    text = element_text(size = 16, colour = "black"),
    panel.grid.major = element_line(colour = "grey99"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    aspect.ratio = 1,
  )+
  xlab("PC1 (17%)")+
  ylab("PC2 (11%)")

```

Finally, lets see if the groups label correlates with microbiome diversity.  

```{r adonis}
set.seed(731) 

seq_rel.adonis2 <- adonis2(seqtab.nochim.rel ~  tissue * factor(sex),
                                permutations = 5000, 
                                by = "terms",
                                data = metadata,
                                na.action = na.omit)


seq_rclr.adonis2 <- adonis2(seqtab.nochim ~  tissue * factor(sex) ,
                                method ="robust.aitchison",
                                permutations = 5000, 
                                by = "terms",
                                data = metadata,
                                na.action = na.omit)


seq_rel.adonis2
seq_rclr.adonis2
```


```{r beta_dispersion}
disp.rclr <- betadisper(vegdist(seqtab.nochim, method = "robust.aitchison"), group = metadata$tissue)

disp.rel <- betadisper(vegdist(seqtab.nochim, method = "bray"), group = metadata$tissue)


anova(disp.rclr)
anova(disp.rel)

disp.rclr <- betadisper(vegdist(seqtab.nochim, method = "robust.aitchison"), group = metadata$sex)

disp.rel <- betadisper(vegdist(seqtab.nochim, method = "bray"), group = metadata$sex)


anova(disp.rclr)
anova(disp.rel)

```

```{r cca_plot}

seqtab.nochim.rel.cca <- cca(seqtab.nochim.rel ~ sex * tissue, data = metadata)

set.seed(731)
asv_cca.anova <-anova(seqtab.nochim.rel.cca, by ="terms")

cca.df <- data.frame(scores(seqtab.nochim.rel.cca, tidy = F, display = "wa", choices = c(1:7)))

cca.df$tissue <- metadata$tissue
cca.df$sex <- metadata$sex

cca.ggplot <- ggplot(data = cca.df, 
                     aes(x = CCA1,
                         y = CCA2, 
                         color = tissue, 
                         shape = factor(sex))
                    )

png(filename = "../../analysis/figs/sex_biogeo_cca.png", res = 300, width = 7,height = 7, units = "in")
cca.ggplot +
  geom_point(size = 8, alpha = .7)+
  scale_color_brewer("Tissue", palette = "Dark2", labels = c("Proximal", "Middle", "Distal", "Feces"))+
  scale_shape_manual("Sex", values = c(16, 17), labels = c("Male", "Female"))+
  theme(
    text = element_text(size = 20, colour = "black"),
    panel.grid.major = element_line(colour = "grey99"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    aspect.ratio = 1, legend.position = "bottom", legend.box="vertical"
  )

dev.off()

cca_rev.ggplot <-  ggplot(data = cca.df, 
                     aes(x = CCA1,
                         y = CCA2, 
                         shape = tissue, 
                         fill = factor(sex)
                         )
)

png(filename = "../../analysis/figs/sex_biogeo_cca_rev.png", res = 300, width = 7,height = 7, units = "in")

cca_rev.ggplot +
  geom_point(size = 8, alpha = .7)+
  scale_fill_brewer("Sex", palette = "Dark2", labels =  c("Female", "Male"))+
  scale_shape_manual("Tissue", values = c(21, 22, 23,25), labels = c("Proximal", "Middle", "Distal", "Feces"))+
  theme(
    text = element_text(size = 20, colour = "black"),
    panel.grid.major = element_line(colour = "grey99"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    aspect.ratio = 1,
    legend.position = "bottom", 
    legend.box="vertical"
  )+
  guides(fill = guide_legend(
    override.aes=list(shape = 21)), 
    shape =  guide_legend(
    override.aes=list(fill = "grey20")))

dev.off()
```

#### Statistical analysis of sex

I prefer to use counts data (rarefied) for regression with MB data. We will use a model comparison procedure to control false discovery rate. We will begin with feces.  
```{r stats_fecal}

# make a df with just fecal samples

#sample names to keep
fecal.names <- rownames(metadata[which(metadata$tissue == "F"),])

#make meta
fecal.metadata <- metadata[which(rownames(metadata) %in% fecal.names),]

#make asv table
fecal.rare <- seqtab.nochim.rare[which(rownames(seqtab.nochim.rare) %in% fecal.names),]
dim(fecal.rare)

fecal.rare <- fecal.rare[,which(colSums(fecal.rare) > 0)]
dim(fecal.rare)

fecal.stats.obj <- NULL 

for(i in 1:ncol(fecal.rare)){
  fit0 <- NULL
  fit1 <- NULL 
  fit.anova <- NULL
  
  try(fit0 <- glm.nb(fecal.rare[,i] ~ 1))
  try(fit1 <- glm.nb(fecal.rare[,i] ~ sex , data = fecal.metadata))
  try(fit.anova <- anova(fit0, fit1))
  
  fecal.stats.obj[[colnames(fecal.rare)[i]]]$null <- fit0
  fecal.stats.obj[[colnames(fecal.rare)[i]]]$alt <- fit1
  fecal.stats.obj[[colnames(fecal.rare)[i]]]$fit.aov <- fit.anova
}

#make a summary data frame
fecal.stats.df <- data.frame(asv = names(fecal.stats.obj),
                             p.val = sapply(fecal.stats.obj, FUN = function(x){x$fit.aov$`Pr(Chi)`[2]}),
                             fdr = p.adjust(sapply(fecal.stats.obj, FUN = function(x){x$fit.aov$`Pr(Chi)`[2]}), method = "fdr")

                             
                             )

fecal.sig <- fecal.stats.df[which(fecal.stats.df$fdr < .2),"asv"]

for(i in fecal.sig){
  boxplot(fecal.rare[,i] ~ sex , data = fecal.metadata, main = i)
}

```


```{r stats_pg}

# make a df with just fecal samples

#sample names to keep
pg.names <- rownames(metadata[which(metadata$tissue == "PG"),])

#make meta
pg.metadata <- metadata[which(rownames(metadata) %in% pg.names),]

#make asv table
pg.rare <- seqtab.nochim.rare[which(rownames(seqtab.nochim.rare) %in% pg.names),]
dim(pg.rare)

pg.rare <- pg.rare[,which(colSums(pg.rare) > 0)]
dim(pg.rare)

pg.stats.obj <- NULL 

for(i in 1:ncol(pg.rare)){
  fit0 <- NULL
  fit1 <- NULL 
  fit.anova <- NULL
  
  try(fit0 <- glm.nb(pg.rare[,i] ~ 1))
  try(fit1 <- glm.nb(pg.rare[,i] ~ sex , data = pg.metadata))
  try(fit.anova <- anova(fit0, fit1))
  
  pg.stats.obj[[colnames(pg.rare)[i]]]$null <- fit0
  pg.stats.obj[[colnames(pg.rare)[i]]]$alt <- fit1
  pg.stats.obj[[colnames(pg.rare)[i]]]$fit.aov <- fit.anova
}

#make a summary data frame
pg.stats.df <- data.frame(asv = names(pg.stats.obj),
                             p.val = sapply(pg.stats.obj, FUN = function(x){x$fit.aov$`Pr(Chi)`[2]}),
                             fdr = p.adjust(sapply(pg.stats.obj, FUN = function(x){x$fit.aov$`Pr(Chi)`[2]}), method = "fdr")

                             
                             )

pg.sig <- pg.stats.df[which(pg.stats.df$fdr < .2),"asv"]

for(i in pg.sig){
  boxplot(pg.rare[,i] ~ sex , data = pg.metadata, main = i)
}



```
